---
title: "EDA Codebook"
author: "Sahithi R. Gangaram, Sangavi Manickavel, Tiger Liao"
date: "2024-07-22"
output:
  html_document: default
  pdf_document: default
---

---
title: "EDA Codebook"
author: "Sahithi R. Gangaram, Sangavi Manickavel, Tiger Liao"
date: "2024-07-22"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load necessary libraries
if (!requireNamespace("usmap", quietly = TRUE)) install.packages("usmap") 
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
if (!requireNamespace("scales", quietly = TRUE)) install.packages("scales")
if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")
if (!requireNamespace("plotly", quietly = TRUE)) install.packages("plotly")

library(dplyr)
library(ggplot2)
library(usmap)
library(scales)
library(tidyr)
library(plotly)

# Load dataset
load("C:/Users/Owner/Desktop/EDA Project/N-MHSS-2010-DS0001-data/N-MHSS-2010-DS0001-data-r.rda")

dataset <- da34945.0001

```

# Introduction

This analysis explores trends in mental health treatment access in the United States using the National Mental Health Services Survey (N-MHSS) data from 2010. The aim is to investigate the geographical disparities in access to mental health treatment facilities and the distribution of different types of facilities, such as non-profit, public, and outpatient centers, across the US. Additionally, this study will examine whether there are differences in the utilization of mental health services based on the type of facility. By understanding these patterns, we hope to identify areas with limited access to mental health services and potential disparities in service utilization, which can inform policy and resource allocation decisions.

# Exploratory Objectives and Their Formulation
# Objective 1: Which states have the most and fewest mental health treatment facilities?

**Formulation and Motivation:**
The motivation behind this question is to identify geographical disparities in the availability of mental health treatment facilities. By understanding which states have the most and fewest facilities, we can identify regions that may need increased resources and support for mental health services. This insight can inform policy decisions and resource allocation to ensure equitable access to mental health care across the United States.

# Objective 2: What are the most common mental health services offered across different states?

**Formulation and Motivation:**
This question aims to identify the core services that mental health facilities provide and understand their distribution across states. Knowing the most common services helps in evaluating the comprehensiveness of mental health care provided. It also allows us to identify gaps in service provision, enabling stakeholders to address unmet needs and enhance the overall quality of mental health care.

# Objective 3: What are the differences in the average number of male and female outpatients treated at mental health facilities, and how do these numbers vary by state?

**Formulation and Motivation:**
Exploring gender differences in outpatient treatment helps us understand whether there are disparities in access to mental health care between men and women. By examining these differences on a state level, we can identify states where these disparities are most pronounced. This information can guide targeted interventions to ensure that both men and women have equitable access to mental health services.

# Objective 4: How many facilities offer specialized programs for different age groups, and what is the distribution of these facilities across states?

**Formulation and Motivation:**
This question is designed to investigate the availability of age-specific mental health programs, such as those for youth, adults, and seniors. Understanding the distribution of these specialized programs helps in identifying whether certain age groups are underserved. This information is crucial for developing age-appropriate mental health services and ensuring that facilities cater to the unique needs of different age populations.

# Objective 5: What is the most common kind of facility? What kinds of facilities offer the widest variety of services?

**Formulation and Motivation:** 

The goal here is to understand the types of facilities that dominate the mental health care landscape and which ones provide the most comprehensive care. By identifying the most common facility types and those offering the widest range of services, we can evaluate the overall structure and capacity of the mental health care system. This insight can help in optimizing facility types to meet the diverse needs of the population.

# What Led to These Questions:
The formulation of these questions is driven by a need to comprehensively understand the mental health care infrastructure in the United States. Recognizing disparities in facility distribution, service provision, and access to care across different demographics is essential for improving mental health outcomes. By addressing these questions, we aim to highlight areas requiring attention, guide resource allocation, and inform policy decisions to build a more equitable and effective mental health care system.

# Data Exploration
The following steps outline our approach to exploring the N-MHSS dataset to address our research objectives.

# Reading and Cleaning Data
First, we load and prepare the dataset. This involves renaming unclear variables and transforming data for clarity. Below are the variables we are particularly interested in for our analysis.

```{r}
# Extract variable names and descriptions
variable_names <- colnames(dataset)

variable_descriptions <- sapply(variable_names, function(var) {
  attr <- attr(dataset[[var]], "label")
  if (is.null(attr)) {
    return("Description not available")
  } else {
    return(attr)
  }
})

# Create a dataframe with variable names and descriptions
variables_df <- data.frame(
  Variable = variable_names,
  Description = variable_descriptions
)

# Clean the dataset
cleaned_dataset <- dataset %>%
  mutate(state_name = gsub("^[^ ]* ", "", as.character(STFIPS))) %>%
  rename(
    state_fips = STFIPS,
    facility_type = FACILITYTYPE,
    intake_service = MHINTAKE,
    diag_eval_service = MHDIAGEVAL,
    substance_abuse_service = SASERV,
    mental_health_service = MENTALHTHSERV,
    male_outpatients = OPSEXTOTM,
    female_outpatients = OPSEXTOTF,
    youth_programs = CHILDADOL,
    adult_programs = ADULTS,
    senior_programs = SENIORS
  )

# Define variable descriptions as a named vector
variable_descriptions <- c(
  STFIPS = "Census State FIPS codes label",
  FACILITYTYPE = "Facility type",
  MHINTAKE = "QA1_1 Facility offers intake services",
  MHDIAGEVAL = "QA1_2 Facility offers diagnostic evaluation",
  SASERV = "QA1_4 Facility offers substance abuse treatment services",
  MENTALHTHSERV = "QA1_5 Facility offers mental health treatment services",
  OPSEXTOTM = "QB5b_num_mal Male outpatients - Total",
  OPSEXPERM = "QB5b_per_mal Male outpatients - Percent",
  OPSEXTOTF = "QB5b_num_fem Female outpatients - Total",
  OPSEXPERF = "QB5b_per_fem Female outpatients - Percent",
  CHILDADOL = "QA12_1 Youth accepted at facility",
  ADULTS = "QA12_2 Adults accepted at facility",
  SENIORS = "QA12_3 Seniors accepted at facility"
)

variables_df <- data.frame(
  Variable = names(variable_descriptions),
  Description = unname(variable_descriptions)
)

# Display the dataframe
knitr::kable(variables_df, caption = "Variables and Descriptions")

```
This code chunk succinctly prepares the dataset for analysis by renaming and transforming the relevant variables. The displayed table lists the key variables and their descriptions, which are crucial for understanding trends in mental health treatment access, service provision, and demographic disparities across the United States.

# Analysis and Visualizations
# Number of Facilities by State
To address the first objective, we visualize the number of mental health facilities by state. This helps us identify which states have the highest and lowest number of facilities.

```{r}
# Count the number of facilities by state
state_facility_counts <- cleaned_dataset %>%
  count(state_name) %>%
  rename(num_facilities = n)

# Plot the number of facilities by state
ggplot(state_facility_counts, aes(x = reorder(state_name, -num_facilities), y = num_facilities)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Number of Mental Health Facilities by State", x = "State", y = "Number of Facilities") +
  theme_minimal() +
  theme(axis.text = element_text(size = 7))
```

# Most Common Mental Health Services
To answer the second objective, we analyze the frequency of different mental health services offered across facilities.

```{r}
# Ensure that the service columns are numeric
cleaned_dataset <- cleaned_dataset %>%
  mutate(
    intake_service = as.numeric(intake_service),
    diag_eval_service = as.numeric(diag_eval_service),
    substance_abuse_service = as.numeric(substance_abuse_service),
    mental_health_service = as.numeric(mental_health_service)
  )

# Count the occurrences of each service
services_count <- cleaned_dataset %>%
  summarise(
    intake_service = sum(intake_service, na.rm = TRUE),
    diag_eval_service = sum(diag_eval_service, na.rm = TRUE),
    substance_abuse_service = sum(substance_abuse_service, na.rm = TRUE),
    mental_health_service = sum(mental_health_service, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = everything(), names_to = "service", values_to = "count")

# Plot mental health services
ggplot(services_count, aes(x = reorder(service, -count), y = count, fill = service)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Most Common Mental Health Services Offered", x = "Service", y = "Count of Facilities") +
  scale_y_continuous(breaks = seq(0, max(services_count$count), by = 5000)) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 12),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

```
# Average Number of Male and Female Outpatients
For the third objective, we calculate the average number of male and female outpatients treated at mental health facilities and visualize the state-wise differences.

```{r}
# Convert outpatient variables to numeric
cleaned_dataset <- cleaned_dataset %>%
  mutate(
    male_outpatients = as.numeric(male_outpatients),
    female_outpatients = as.numeric(female_outpatients)
  )

# Summarize the average number of male and female outpatients by state
outpatients_avg <- cleaned_dataset %>%
  group_by(state_name) %>%
  summarise(
    avg_male_outpatients = mean(male_outpatients, na.rm = TRUE),
    avg_female_outpatients = mean(female_outpatients, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(avg_male_outpatients, avg_female_outpatients), names_to = "gender", values_to = "avg_outpatients")

# Plot the average number of male and female outpatients by state
ggplot(outpatients_avg, aes(x = reorder(state_name, avg_outpatients), y = avg_outpatients, fill = gender)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap(~ gender, scales = "free_y") +
  labs(
    title = "Average Number of Male and Female Outpatients Treated by State",
    x = "State",
    y = "Average Number of Outpatients",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 4),
    axis.title = element_text(size = 8),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "none"
  )

```

# Distribution of Specialized Programs
We analyze the distribution of specialized programs for different age groups across states and visualize the findings on a map.

# Data Preparation
We start by ensuring that the relevant columns are converted to numeric values for summing.
```{r}
library(dplyr)
library(tibble)
library(formattable)

# Ensure the relevant columns are factors and convert them to numeric for summing
cleaned_dataset <- cleaned_dataset %>%
  mutate(
    youth_programs = ifelse(youth_programs == "(1) Yes", 1, 0),
    adult_programs = ifelse(adult_programs == "(1) Yes", 1, 0),
    senior_programs = ifelse(senior_programs == "(1) Yes", 1, 0)
  )
```

# Summary of Specialized Programs
We create a summary dataset to count the number of specialized programs offered by each state.
```{r}
# Create a summary dataset for specialized programs, counting the "yes" values
specialized_programs <- cleaned_dataset %>%
  select(state_fips, youth_programs, adult_programs, senior_programs) %>%
  group_by(state_fips) %>%
  summarise(
    youth_programs_count = sum(youth_programs, na.rm = TRUE),
    adult_programs_count = sum(adult_programs, na.rm = TRUE),
    senior_programs_count = sum(senior_programs, na.rm = TRUE)
  )

# Convert the summary dataset to a tibble
specialized_programs_tibble <- as_tibble(specialized_programs)
```

# Highlighting Maximum Counts
We identify the states with the highest counts for each type of program and create a formattable table to highlight these states.
```{r}
# Find the largest number in each column
max_youth <- max(specialized_programs_tibble$youth_programs_count, na.rm = TRUE)
max_adult <- max(specialized_programs_tibble$adult_programs_count, na.rm = TRUE)
max_senior <- max(specialized_programs_tibble$senior_programs_count, na.rm = TRUE)

# Create a formattable table with the largest number in bold
formatted_table <- formattable(specialized_programs_tibble, list(
  youth_programs_count = formatter("span",
                                   style = x ~ ifelse(x == max_youth, "font-weight: bold;", "font-weight: normal;")),
  adult_programs_count = formatter("span",
                                   style = x ~ ifelse(x == max_adult, "font-weight: bold;", "font-weight: normal;")),
  senior_programs_count = formatter("span",
                                    style = x ~ ifelse(x == max_senior, "font-weight: bold;", "font-weight: normal;"))
))

# Print the formatted table
formatted_table
```

# Visualization of Specialized Programs
We then visualize the distribution of these specialized programs across states using maps.
```{r}
# Convert state_fips to numeric
specialized_programs$state_fips <- as.numeric(gsub("\\D", "", specialized_programs$state_fips))

# Prepare the state map data
state_map <- map_data("state")
state_fips <- data.frame(state.abb = state.abb, region = tolower(state.name), state_fips = as.numeric(fips(state.abb)))

# Join the specialized programs data with state map data
specialized_programs_map <- state_map %>%
  left_join(state_fips, by = "region") %>%
  left_join(specialized_programs, by = "state_fips")
```

# Distribution of Youth Programs
```{r}
# Plot the distribution of youth programs across states
ggplot(specialized_programs_map, aes(x = long, y = lat, group = group, fill = youth_programs_count)) +
  geom_polygon(color = "white") +
  scale_fill_continuous(name = "Youth Programs", low = "white", high = "blue", na.value = "grey50") +
  labs(title = "Distribution of Youth Programs Across States") +
  theme_minimal() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank())
```

# Distribution of Adult Programs
```{r}
# Plot the distribution of adult programs across states with a darker green
ggplot(specialized_programs_map, aes(x = long, y = lat, group = group, fill = adult_programs_count)) +
  geom_polygon(color = "white") +
  scale_fill_continuous(name = "Adult Programs", low = "white", high = "#006400", na.value = "grey50") +
  labs(title = "Distribution of Adult Programs Across States") +
  theme_minimal() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank())
```

# Distribution of Senior Programs
```{r}
# Plot the distribution of senior programs across states
ggplot(specialized_programs_map, aes(x = long, y = lat, group = group, fill = senior_programs_count)) +
  geom_polygon(color = "white") +
  scale_fill_continuous(name = "Senior Programs", low = "white", high = "red", na.value = "grey50") +
  labs(title = "Distribution of Senior Programs Across States") +
  theme_minimal() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank())
```

# Explanation:
These visualizations provide insight into the availability of specialized programs for different age groups across the United States. By examining these maps, policymakers and healthcare providers can identify states with insufficient specialized programs and work towards improving access to age-specific mental health services.

## Number of Services Offered by Each Facility Type
Next, we analyze the number of services offered by each type of mental health facility to understand which facility types provide the most comprehensive care.

# Data Preparation
We prepare the data by summarizing the number of services offered by each facility type.
```{r}
# Summarize the number of services offered by each facility type
services_offered_by_facility <- cleaned_dataset %>%
  select(facility_type, intake_service, diag_eval_service, substance_abuse_service, mental_health_service) %>%
  pivot_longer(cols = intake_service:mental_health_service, names_to = "service", values_to = "offered") %>%
  group_by(facility_type) %>%
  summarise(num_services_offered = sum(offered, na.rm = TRUE)) %>%
  arrange(desc(num_services_offered))

```

# Shortening Labels for Better Visualization
We shorten the facility type labels to improve the readability of the plot.

```{r}
# Shorten the facility type labels for better fit
services_offered_by_facility <- services_offered_by_facility %>%
  mutate(facility_type = case_when(
    facility_type == "(5) Outpatient, day treatment or partial hospitalization mental health facility" ~ "Outpatient/Partial Hosp.",
    facility_type == "(6) Multi-setting (non-hospital) mental health facility" ~ "Multi-setting",
    facility_type == "(1) Psychiatric Hospital" ~ "Psychiatric Hosp.",
    facility_type == "(4) Residential treatment center for adults" ~ "Residential (adults)",
    facility_type == "(3) Residential treatment center for children" ~ "Residential (children)",
    facility_type == "(2) Separate inpatient psychiatric unit of a general hospital" ~ "Inpatient Unit"
  ))

```

# Visualization of Services Offered
We create an interactive bar chart to show the number of services offered by each facility type.
```{r}
# Create the ggplot object
p <- ggplot(services_offered_by_facility, aes(x = reorder(facility_type, -num_services_offered), y = num_services_offered, text = paste("Facility Type: ", facility_type, "<br>Number of Services: ", num_services_offered))) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Number of Services Offered by Each Facility Type", x = "Facility Type", y = "Number of Services Offered") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 10),
    legend.position = "none"
  )

# Convert the ggplot object to a plotly object for interactivity
p <- ggplotly(p, tooltip = "text")

# Add annotations to explain the importance of the information
p <- p %>% layout(
  annotations = list(
    list(
      x = 1,
      y = 1.05,
      showarrow = FALSE,
      xref = 'paper',
      yref = 'paper',
      xanchor = 'right',
      yanchor = 'bottom',
      align = 'left',
      font = list(size = 12)
    )
  )
)

# Display the interactive plot
p
```
# Explanation:
1. *Data Preparation:* 

The service columns (intake_service, diag_eval_service, substance_abuse_service, and mental_health_service) are converted to numeric and summarized by facility type.

2. *Plotting:*

* A bar chart is created using ggplot2 to visualize the number of services offered by each facility type.
* The height of the bars represents the number of services provided by each facility type.
* `coord_flip` is used to flip the axes for better readability.

## Analysis

The initial visualizations provide a comprehensive overview of the geographical distribution and the types of services offered by mental health facilities across different states.

1. **Geographical Distribution of Facilities:**
* The number of mental health facilities varies significantly across states, with California, New York, and Pennsylvania having the highest number of facilities, while states like North Dakota, Delaware, and Hawaii have the fewest.

* This distribution indicates potential disparities in access to mental health care, suggesting that residents in states with fewer facilities might face challenges in obtaining timely and adequate mental health services.

2. **Common Mental Health Services:**
* Diagnostic evaluation services and intake services are the most commonly offered services across mental health facilities, followed by substance abuse services and mental health treatment services.
* This suggests that most facilities are equipped to perform initial assessments and evaluations, which are critical for identifying the appropriate treatment plans for patients.

3. **Average Number of Male and Female Outpatients:**
* The analysis of outpatients treated at mental health facilities reveals notable gender disparities.
* States like Tennessee, South Carolina, and Arizona have high average numbers of both male and female outpatients, indicating a higher utilization of mental health services in these states.

4. **Distribution of Specialized Programs:**
* Youth, adult, and senior programs are unevenly distributed across states.
* States like California and New York offer a higher number of specialized programs, while many states in the Midwest and the South have fewer such programs.

5. **Number of Services Offered by Facility Type:**
* Outpatient/Partial Hospitalization facilities offer the highest number of services, indicating their role in providing comprehensive mental health care.
*Other facility types, such as psychiatric hospitals and residential treatment centers, offer fewer services, which might reflect their specialized focus areas.

## Looking Forward

The initial analysis suggests several areas for further investigation and action:

1. *Addressing Geographical Disparities:*
* Policy interventions are needed to address the uneven distribution of mental health facilities and services. States with fewer facilities should be prioritized for resource allocation and support to ensure equitable access to mental health care.

2. *Understanding Gender Disparities:*
* Further research is required to understand the underlying causes of gender disparities in mental health service utilization. This can help in designing targeted interventions to ensure both men and women have equal access to mental health services.

3. *Expanding Specialized Programs:*
* Increasing the availability of specialized programs for different age groups is essential. States with fewer such programs should be targeted for the development and implementation of youth, adult, and senior mental health services.

4. *Evaluating Facility Types and Services:*
* Analyzing which facility types provide the most comprehensive care can inform decisions about expanding or modifying existing facilities. Outpatient/Partial Hospitalization facilities, which offer a wide range of services, can serve as a model for developing comprehensive care centers in other facility types.

5. *Exploring Intersectionality:*
* Investigating the intersectionality of age, gender, and race in accessing mental health care can reveal specific barriers faced by different demographic groups. This information is crucial for developing inclusive and effective mental health policies.

6. *Aligning Services with Population Needs:*
* Assessing whether the prevalence of certain facility focuses, such as substance abuse treatment, aligns with the actual needs of the population can ensure that mental health services are appropriately tailored to address the most pressing issues.

By addressing these areas, we can work towards a more equitable and effective mental health care system that meets the diverse needs of the population.